---
title: "GASImpactModel"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GASImpactModel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(GASImpactModel)
```


## Introduction
A static cohort model was developed to estimate the projected health impact of 
group A Streptococcus (GAS) vaccination using country-specific demographic data. 
The model is available as an R package GASImpactModel 
(https://github.com/fionagi/GASImpactModel) and a user-friendly (R Shiny) web application (https://github.com/fionagi/GASImpactModel_App). Vaccination impact is estimated 
in terms of reduction in the burden of several major GAS disease states and sequelae. 
Burden estimation comprised of episodes of acute GAS disease (pharyngitis, impetigo, 
invasive disease, and cellulitis) and cases of rheumatic heart disease (RHD), 
deaths due to severe GAS disease (invasive disease and RHD), and disability-adjusted 
life years (DALYs) due to each GAS disease. The reduction in disease burden is in 
direct proportion to vaccine efficacy, vaccine coverage, and vaccine-derived immunity 
(based on duration of protection and waning dynamics). Through the web application, 
impact metrics for a selected vaccination scenario can be visualised for any of 205 
countries and regions. The app shows the predicted lifetime health benefits, from age of vaccination,
associated with the vaccination of multiple cohorts during the period selected. 
Direct effects of vaccination are included, and indirect (herd) effects are excluded; 
therefore, the estimated health benefits of GAS vaccination are conservative 
(if GAS vaccination prevents population transmission).


### Functions
The main GASImpactModel function is:

  * runModel

Other functions include:

  * getConditionData
  * getInitPop
  * getPopData
  * getMorData
  * getLifeExData
  * immunity
  * preVaxxModel
  * propVModel
  * findDalys
  
  * makeBarPlot
  * makePlot
  * makePlotAvert
  
R code for the functions are in the R-folder of the package, i.e. ./R/

Documentation of a function can be viewed, by running ?functionname in the console, i.e. ?getConditionData
```{r eval=FALSE}
# View documentation of getConditionData function
?getConditionData

```

The source code of a function can be viewed by running the functionname without any brackets in the console.
```{r eval=FALSE}
# View source code of getConditionData function
getConditionData

```


### Data
Several data sets are included in the package. The data files (.rda files) are stored
in the data folder of the package, i.e. ./data/

  * data.region
  * data.cellulitis2019
  * data.rhd2019
  * data.popbyage.1950_to_2020
  * data.popbyage.pred
  * data.mortality.1950_to_2020
  * data.mortality.pred2050
  * data.mortality.pred2100
  * data.lifeEx.1950_to_2020
  * data.lifeEx.pred2100
  * data.coveragebycountry

Documentation of a data set can be viewed, by running ?dataset-name in the console, i.e. ?data.cellulitis2019
```{r eval=FALSE}
# View documentation of ?data.cellulitis2019 data set
?data.cellulitis2019

```

A data set can be viewed by typing the data file name in the console.
```{r eval=FALSE}
# View entire data set of incidence, death rates and DALYs for all countries for cellulitis
data.cellulitis2019

```


### Examples
The "get" functions (e.g. getConditionData, getInitPop, getMorData, etc.) can
be used to retrieve country specific data from the data sets. For example, getConditionData 
can be used to retrieve cellulitis incidence data for Australia 

```{r}
getConditionData(location = "Australia", condition = "Cellulitis", metric = "Rate")[[1]]

```

If the condition is not cellulitis or RHD, the incidence rates
will not be country-specific. Only for cellulitis and RHD does the
metric need to be defined, and a list (rather than just a data frame) containing 
incidence (first list item), deaths (second list item) and DALYs (third list item) 
will be returned as this is the full data provided by the Institute for Health 
Metrics and Evaluation (IHME) and stored as a data set in the R package. 

Another example; use the getInitPop function to find the population at vaccination
age assuming a vaccine is introduced in Australia in 2022, considering 10 cohorts
from year of vaccination with a vaccination age of 5.

```{r}
getInitPop(location = "Australia", yearV = 2022, pYears = 10, ageV = 5, birth = FALSE)
```

Note: If you want to find the birth population for these same cohorts, then "birth = TRUE". 

To model the impact of vaccination on a particular population, for a GAS disease, 
the R Shiny app is a user-friendly way of defining a vaccination scenario and
visualising the model results. The GASImpactModel_App uses the R package 
GASImpactModel. 

To access the app for modelling: 

1.	Make sure you have the “shiny” package installed and loaded: 
```{r eval=FALSE}
install.packages("shiny")
library(shiny)
```

2.	Then run app code directly from the GitHub repository in your R console by 
using the following command:
```{r eval=FALSE}
runGitHub("GASImpactModel_App", "fionagi")
```


To run the model using only the R package, we first need to retrieve the
incidence, all-cause mortality and population data for the scenario of interest.

In this example, we are modelling invasive infection, in Australia, with 2022
the year of vaccination, age of vaccination 0 (maximum age of cohort is 99), for
10 years after year of vaccination (so 11 years in total). The parameter "impType",
refers to how the metrics are estimated - either by year of birth, year of vaccination
or calendar year. In this example, the results will be presented considering the
impact of a particular year’s vaccination activity, "Year of vaccination".
```{r}
#get incidence rates
incR <- getConditionData(location = "Australia", condition = "Invasive infection")
#get all-cause mortality rates for each cohort
mProb <- getMorData(location = "Australia", yearV = 2022, pYears = 10,
                        ageV = 0, maxAge = 99, impType = "Year of vaccination")
#get population at year of vaccination for each cohort
initPop <- getInitPop(location = "Australia", yearV = 2022, pYears = 10, 
                        ageV = 0, birth = FALSE)
    
```


Once we have the incidence data, we can plot the number of cases per 100 persons, by age group.
This data is the constant (over projected years), pre-vaccination assumed incidence rate.
```{r fig1, fig.height = 5, fig.width = 11}
#plot incidence data as number of cases per 100 persons
incR_per100 <- incR
incR_per100[,"val"] <- 100*incR_per100[,"val"]
makeBarPlot(incR_per100, ylabel = "Number of cases per 100 persons", colFill = "steelblue")
```
Now that we have the required data, we can model the impact of a particular vaccination scenario.
In this example we consider a vaccine efficacy of 70%, vaccine coverage of 50%,
a durability of 10 years with no waning of immunity and no linear "ramp" up to
maximum vaccine coverage.

Note: rate here corresponds to the incidence data. For RHD and cellultitis the values
are per 100,000 persons. For all other conditions, it is per person hence rate = 1.

The function runModel takes the input data and vaccination parameter values and returns a list where
each item in the list is a data frame. The list gives us the pre-vaccination and vaccination
incident counts, DALYS and deaths (if relevant), as well
as the modelled population.

```{r}
#vaccine efficacy as %
efficacy <- 70
#vaccine coverage as %
coverage <- 50
#overall vaccine effect, incorporating vaccine efficacy and coverage
overallEff <- (efficacy*coverage)/100

#run model
impModels <- runModel(location = "Australia", condition = "Invasive infection", 
                        inc = incR, rate = 1, mortality = mProb, yearV = 2022,
                           vaccAge = 0, maxAge = 99, vaccEff = overallEff,
                           vaccDur = 10, waning = FALSE, ramp = FALSE,
                           impType = "Year of vaccination", pYears = 10,
                           initPop = initPop)

#save list results returned from "runModel"
noVacc_counts <- impModels[[1]]
vacc_counts <- impModels[[2]]
noVacc_dalys <- impModels[[3]]
vacc_dalys <- impModels[[4]]
noVacc_deaths <- impModels[[5]]
vacc_deaths <- impModels[[6]]
pop <- impModels[[7]]

```

The results in terms of incidents, DALYs and deaths averted can then be plotted.
```{r fig2, fig.height = 9, fig.width = 6}
#plot averted results
p_countsA <- makePlotAvert(noVacc_counts, vacc_counts, ylabel = "Number of cases",
                         vAge = 0, vYear = 2022, impType = "Year of vaccination",
                         pYears = 10)

p_dalysA <- makePlotAvert(noVacc_dalys, vacc_dalys, ylabel = "DALYs",
                          vAge = 0, vYear = 2022, impType = "Year of vaccination",
                          pYears = 10)

p_deathsA <- makePlotAvert(noVacc_deaths, vacc_deaths, ylabel = "Deaths",
                           vAge = 0, vYear = 2022, impType = "Year of vaccination",
                           pYears = 10)
ggpubr::ggarrange(p_countsA, p_dalysA, p_deathsA, ncol = 1, nrow = 3)

```


